\LARGE \bf Human-Robot Interaction: Tackling the AI Challenges

### 2014-05-16 paper.tex
Rachid's comments (typos/style...)

Human-Robot interaction is an area filled with challenges for artificial
intelligence: dynamic, partially unknown environments that are not originally
designed for autonomous machines; a large variety of situations and objects to
deal with, with possibly complex semantics; physical interactions with humans
that requires fine, low-latency control, representation and management of
several mental models, good situation assessment skills...the list goes on.

This article sheds light on some key decisional issues that are to be tackled
for a cognitive robot to share space and tasks with a human, and present our
take on these challenges. We adopt a constructive approach based on the
identification and the effective implementation of individual and collaborative
skills. These cognitive abilities cover geometric reasoning and situation
assessment mainly based on perspective-taking and affordances, management and
exploitation of each agent (human and robot) knowledge in separate cognitive
models, natural multi-modal communication, human-aware task planning, and human
and robot interleaved plan achievement.

@@

Human-robot interaction is a challenge for artificial intelligence. This field
lays at the crossroad of several domains of AI and requires to tackle them in a
holistic manner: Modelling humans and human cognition; acquiring, representing,
manipulating in a tractable way abstract knowledge at the human level; reasoning
on this knowledge to make decisions; and eventually instantiating those
decisions into physical actions both legible to and in coordination with humans.
Many AI techniques are invited, from visual processing to symbolic reasoning,
from task planning to \emph{theory of mind} building, from reactive control to
action recognition and learning.

@@

We focus on a specific class of interactions: collaborative task achievement
supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
illustrates this context: the two agents share a common space and exchange
information through multiple modalities, and the robot is expected to achieve
interactive object manipulation, fetch and carry tasks and other similar
domestic tasks by taking into account, at every stages, the intentions, beliefs,
perspectives, skills of the human partner.  Namely, the robot must be able to
recognise, understand and participate to communication situations, both explicit
(\eg the human addresses verbally the robot) and implicit (\eg the human points
to an object); the robot must be able to take part to joint actions, both
pro-actively (by planning and proposing resulting plans to the human) and
reactively; the robot must be able to move and act in a safe, efficient and
legible way, taking into account social rules like proxemics.

@@

\emph{Communication}, \emph{Joint Actions}, \emph{Human-Aware Execution}: these
three challenges structure the whole field of human-robot interaction research,
and they can be analysed in terms of the cognitive skills that they require.

@@

The robot controller (with the help of a task planner) decides what action to
execute next, and who should perform it from the human or the robot (or both in
case of a joint action), and finally controls or monitors its execution. The
operation continues until the goal is achieved, is declared unachievable or is
abandoned by the human.

@@

Such abilities must be designed and implemented in generic ways and
must provide several levels of parametrization such as they adapt to
various environments, different tasks and variable levels of engagement of the
robot, ranging from teammate behaviour to assistant or pro-active helper.

@@

The next section introduces the architecture as a whole, as well as the
knowledge model that we have developed for our robots. Then,
section~\ref{sec:impl} gives some insights on each of the main components of the
architecture, and try to highlight their significance for artificial
intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
practical way what can be currently achieved with our robots.
Section~\ref{sect|conclusion} finally restate our main contributions and discuss
the key challenges that human-robot interaction brings to artificial
intelligence.

@@

  %%% Relations between components
  \path (shary.340) edge [<->, bend left] node[label] {motion plan \\ requests} (mhp);
  \path (shary.west) edge [<->, bend right] node[label] {shared \\ plans} (hatp);
  \path (hatp) edge [<->, bend right] node[label] {world model and \\ agents beliefs} (oro.170);
  \path (dialogs) edge [<->, bend left] node[label] {natural language \\ grounding} (oro.190);
  \path (spark.100) edge [->, bend right] node[label] {symbolic \\ facts} (oro);
  \path (shary) edge [<->, bend left] node[label] {events, \\ world model and \\ agents beliefs} (oro);
  \path (shary) edge [<->, bend left] node[label] {action monitoring \\ and management of \\ position hypotheses} (spark);
  \path (lowlevel) edge [->] (spark);
  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] (shary.east);

@@

In particular, dealing with the rich semantics of natural interaction with
humans has been one of our major focus. To this end, we have been deploying
over the last years explicit knowledge representation and manipulation in the
deliberative layer of the robots as a {\it lingua franca} between our
components.

@@

To some extend, this architecture moves away from classical layered approaches
found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
Interactions between components at the deliberative level are mostly
bidirectional and we do not introduce layers of abstraction amongst software
components\footnote{We do have lower-level modules to execute actions or manage
sensors, but all cognition-related modules live in the same global deliberative
layer.}. Dialogue processing, for instance, illustrates this structure. This
component does not purely act as an alternative sensing modality that would be
fed to a distinct deliberative component. On the contrary, it actively queries
the knowledge base to interpret and disambiguate natural language, and in that
regards, it is fully integrated to the deliberative process itself (we detail
natural language processing at section~\ref{sect|com}).

@@

At run-time, the knowledge available to the robot at a given time comes from
three sources. {\it A priori} knowledge is stored in an ontology (the {\sc
OpenRobots} ontology, discussed below) and loaded at start-up. This static source covers the
\emph{common-sense} knowledge of the robot, and optionally some
scenario-specific knowledge (for instance, about objects that are to be
manipulated). The second part of the knowledge is acquired at run-time from
perception, interaction and planning. The next sections detail these processes.
Lastly, the third source of symbolic statements is the reasoner itself.

@@

    \node [node distance=1, below right=of thing] (tthing) {temporal thing} edge (thing);
        \node [below=of tthing] (tte) {thing with temporal extend} edge (tthing);
            \node [below right=of tte] {time interval} edge (tte);
            \node [below=1 of tte] (sit) {situation} edge (tte);
                \node [below right=of sit] (evt) {event} edge (sit);
                    \node [below right=of evt] (act) {action} edge (evt);
                        \node [below=of act] {purposeful action} edge (act);
                \node [below left=of sit] {static situation} edge (sit);

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures to a certain extend that the knowledge stored by the
robot can be shared or extended with well-defined semantics. A good example is
the concept of \emph{Object}: In everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. Normally, a
human is not considered as an object. In {\sc Cyc}, an object has a more precise
definition: it is something \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange informations between artificial
systems in a non-ambiguous way.

@@

The class \concept{PurposefulAction} is the superset of all the actions that are
purposefully performed by the robot (or another agent). Actual actions (\ie
subclasses of \concept{PurposefulAction} like \concept{Give} or
\concept{LookAt}) are not initially asserted in the common-sense ontology. They
are instead added at run-time by the execution controller (in link with the
symbolic task planner) and the natural language processor based on what the
robot is actually able to perform and/or to interpret in the current context
(\ie the current robot configuration and the actions required by the scenario).
The set of actions that the robot can interpret typically corresponds to the
planing domain in use: the robot can only process the \concept{GoTo} action if
the corresponding task description with its pre- and post-conditions is
available to the symbolic task planner.

@@

\emph{Grounding} (also called \emph{anchoring} when specifically referring to
the building of links between \emph{percepts} and \emph{physical
objects}~\cite{Coradeschi2003}) is the task consisting in building and
maintaining a bi-directional link between sub-symbolic representations (sensors
data, low-level actuation) and symbolic representations that can be manipulated
and reasoned about~\cite{Harnad1990}. This represents an important cognitive
skill, in particular in the human-robot interaction context: in that case, the
link that the robot has to established between its percepts and its symbol must
\emph{model} to a large extend the human representations in order to effectively
support communication.

\emph{Symbol grounding} connects hence the knowledge model to the perception and
actuation capabilities of the robot, and our architecture adopts here also a
bidirectional approach.  The different components that we have mentioned so far
provide a \textit{bottom-up} grounding process: geometric reasoning and dialogue
processing modules constantly build and push new symbolic contents about the
world to the knowledge base where it becomes accessible to decisional layers. In
parallel, the knowledge base relies on reasoning in a \textit{top-down} way to
produce new facts, \ie new beliefs that may lead to actual decision making and
physical actions.

@@

\begin{inparaenum} In our context, we call \emph{cognitive skills} the
deliberative behaviours that are \item \emph{stateful} both in terms of
\emph{time} (keeping track of previous states is required for the component
to perform adequately) and \emph{modalities} (for instance, \emph{situated}
dialogue requires interleaving different communication modalities), \item
manipulate \emph{explicit semantics}, typically by the mean of symbolic
reasoning, \item operate at the \emph{human-level}, \ie are legible to the
humans, typically by acting at similar levels of abstraction.\end{inparaenum}

@@

\subsubsection{Reasoning}
\label{reasoning}

@@

We rely however on this mechanism in certain cases: for instance, some modules
like the natural language processor use the {\it short term} memory profile to
mark for a few seconds important concepts that are currently manipulated by the
robot as \emph{active concepts}: If a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature to trace certain knowledge-related processes.

@@

\begin{figure}
\centering
\resizebox{0.8\textwidth}{!}{%
\begin{tikzpicture}[
        >=latex,
        box/.style={draw,rectangle, dotted, minimum width=#1, minimum height=3.8cm},
        box/.default={4.5cm},
        every edge/.style={draw, ultra thick},
        every node/.style={align=center}]

@@

    \node[draw=none] at (0,-1) {\includegraphics[width=0.8\columnwidth]{spark-oro-notext.pdf}};

    \node[box=16cm, minimum height=5.7cm, solid, thick, rounded corners] at (0,1.8) (oro) {};
    \node at (0,4.2) {\sc \Large oro-server};

@@

    \node at (-3,-0.6) {{\it OWL store:} OpenJena};
    \node at (3,-0.6) {{\it Reasoner:} Pellet};

@@

    \node[box=16cm, minimum height=3.5cm, dotted, thick, rounded corners] at (0,-3.75) (spark) {};
    \path (spark.north) edge[->] node[align=left,anchor=west] {Symbolic\\facts} (oro.south);

    \node[anchor=north] at (-5.5,-4.5) {a) Agent-independent\\locations};
    \node[anchor=north] at (-1.8,-4.5) {b) Agent-dependent\\placements};
    \node[anchor=north] at (1.5,-4.5) {c) Visibility};
    \node[anchor=north] at (5.3,-4.5) {d) Reachability};

@@

{\sc Spark} is an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is
updated at run-time by the robot based on its sensors (in most of
our studies, objects are identified and localised through 2D fiducial markers,
while humans are tracked with Kinect-like devices, optionally assisted by motion
capture to accurately track the head motion, which is required to compute what
the human is looking at).

@@

{\sc Spark} is designed to run and continuously update the knowledge base (in
our experiments, it runs at about 10Hz). At each steps, it re-computes spatial
relationships and affordances for the whole scene, and send the delta (new
relations and relations that do not hold anymore) to the knowledge base. This
approach may raise scalability concerns (we did not however hit major
performance issues in our constrained scenarios), but greatly simplifies the
management of the \emph{dynamics} of the knowledge (\emph{When do I discard
outdated knowledge?  When do I update it?}). Since it is equivalent to a reset
of the reasoner domain, it also hides issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).

@@

\item \emph{Sees}: An important ability to know about an agent is to predict
    \emph{What can it see?}, \ie what is within its field of view (FOV). Being
    able to compute this information (by dynamically placing an OpenGL camera at
    the location of the eyes and running occlusion checks from it), the robot
    can reuse it for instance to infers which object a human is searching for
    (the one that is not currently visible, otherwise the user would not be
    searching for it).  In figure~\ref{fig|spark-oro}\emph{c} the field of view
    of a person is illustrated with a grey cone (broader one). While he is able
    to see the two small boxes on the table in front of him, the big box on his
    right is out of his FOV, and therefore, he is not able to see it.

@@

\fxfatal{Rephrase or remove}
It is sometimes difficult or even impossible to see and/or track an object in
certain states. This happens, for instance, when the object is hidden in a
container, when the robot holds it in its gripper, and more generally in any
state in which it is hidden by something else. {\sc Spark} models the possible
symbolic states of objects (whether the object is on a furniture, in an agent
hand, in a container, etc.): According to the robot perception of what has
happened since the object was last seen, the robot builds a belief of the
current possible state with associated probability model.

@@

The main weakness of our current implementation relates to the way semantics are
conveyed: because we use 2D markers for artifacts, we avoid issues related to object
segmentation and recognition. Each object has a unique identifier,
which enable us to load a proper CAD model suitable for 3D spatial reasoning
and prevent recognition mistakes in the knowledge base. While {\sc Spark}
algorithms themselves are ignorant of the input sources, and would work equally
well with a full object recognition stack, we do not have tackled it yet.

Besides, temporal reasoning (essential for accurate action recognition for
instance) is not properly addressed in our stack. Temporal reasoning is used
only locally, and does not allow for tracking of long or global events.

@@

In our experience, natural language processing is one of the fields of
human-robot interaction for which the introduction of the semantic layer has
been most beneficial.  In~\cite{Lemaignan2011a} we detail the techniques and the
tool called {\sc Dialogs} that we have developed for natural English language
parsing and grounding, along with verbalisation and (minimalist) dialogue
management.

@@

The system supports quantification (``give me \{a | the | some | all | any |
...\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
after checking for possible contradictions with existing knowledge),
interprets common temporal and place adverbs (like \emph{above} or
\emph{tomorrow}) and translates to a certain extend \emph{states} (``I'm
tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
hasFeature tired}). We kindly refer the readers to~\cite{Lemaignan2011a} for a
complete discussion of these capabilities and the related algorithms.

\subsubsection{Multi-Modal Communication}

@@

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{human1 desires
?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
behaviour when the human expresses (through dialogue, gestures...) that he
wants to somehow get a toy.

@@

Manipulation and navigation actions listed in table~\ref{table|pyrobots_actions}
rely ona dedicated 3D motion planner ({\sc Mhp}, depicted in
Figure~\ref{fig|archi}): robot placement and end-effectors trajectories are
computed on-line to allow object manipulation that take into account task
specific constraints and human postures, abilities and preferences.
\cite{Sisbot2008, Mainprice2011, Pandey2011} detail these techniques.

@@

This leads to a tight coupling between the symbolic and the geometric realms:
while AI at its origins was mostly a matter of symbolic models, it has been
since recognised that not only the mind is not a purely abstract system,
disconnected from the physical world, but even more, cognition fundamentally
relies on its relation to the physical world (so-called \emph{embodied
cognition}). Varela~\cite{Varela1992} is one of the main discoverer of these
mechanisms, and coined the concept of \emph{enactivism} as the theoretical
framework that study the links between cognition, embodiment and actions.

@@

Grounding also occurs during verbal and non-verbal communication. The {\sc
Dialogs} module grounds new concepts introduced by the speaker by asking questions
until it can attach the new concept to concepts already present in the
knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
it already knows). Non verbal interaction (like gestures) is also partially
grounded: We have presented how pointing, for instance, is used by the robot to
ground \emph{``this''} or \emph{``that''} to the pointed artifact.

@@

As a whole, the components that we have introduced so far build a
knowledge-oriented architecture: Knowledge is explicitly stored in one central
and consistent repository of facts, accessible for all modules. It relies on a
strict formalism (OWL statements), with a well defined vocabulary
(stated in the common-sense ontology). These first two points lead to a
\emph{loosely-coupled architecture} where modules can be removed or replaced
easily by other ones as long as they share the same semantics (modules are
defined by the knowledge they produce).

@@

And finally, this architecture allows for the combination of very different
knowledge modalities in a single homogeneous environment, bringing mutual
benefits to components. For instance, the dialogue processing module can
run without any geometric perception, but its disambiguation routines
can transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).

>>>>

Human-Robot interaction is an area full of challenges for artificial
intelligence: dynamic, partially unknown environments that are not originally
designed for autonomous machines; a large variety of situations and objects to
deal with, with possibly complex semantics; physical interactions with humans
that requires fine, low-latency control, representation and management of
several mental models, pertinent situation assessment skills...the list goes on.

This article sheds light on some key decisional issues that are to be tackled
for a cognitive robot to share space and tasks with a human, and present our
take on these challenges. We adopt a constructive approach based on the
identification and the effective implementation of individual and collaborative
skills. These cognitive abilities cover geometric reasoning and situation
assessment mainly based on perspective-taking and affordances, management and
exploitation of each agent (human and robot) knowledge in separate cognitive
models, natural multi-modal communication, ``human-aware'' task planning, and human
and robot interleaved plan achievement.

@@

Human-robot interaction is a challenge for artificial intelligence. This field
lays at the crossroad of several domains of AI and requires to tackle them in a
holistic manner: modelling humans and human cognition; acquiring, representing,
manipulating in a tractable way abstract knowledge at the human level; reasoning
on this knowledge to make decisions; and eventually instantiating those
decisions into physical actions both legible to and in coordination with humans.
Many AI techniques are invited, from visual processing to symbolic reasoning,
from task planning to \emph{theory of mind} building, from reactive control to
action recognition and learning.

@@

We focus on a specific class of interactions: collaborative task achievement
supported by multi-modal and situated communication. Figure~\ref{fig:hri-dec}
illustrates this context: the two agents share a common space and exchange
information through multiple modalities, and the robot is expected to achieve
interactive object manipulation, fetch and carry tasks and other similar
domestic tasks by taking into account, at every stage, the intentions, beliefs,
perspectives, skills of the human partner.  Namely, the robot must be able to
recognise, understand and participate to communication situations, both explicit
(\eg the human addresses verbally the robot) and implicit (\eg the human points
to an object); the robot must be able to take part to joint actions, both
pro-actively (by planning and proposing resulting plans to the human) and
reactively; the robot must be able to move and act in a safe, efficient and
legible way, taking into account social rules like proxemics.

@@

These three challenges, \emph{Communication}, \emph{Joint Actions},
\emph{Human-Aware Execution}, structure the whole field of human-robot
interaction research, and they can be analysed in terms of the cognitive skills
that they require.

@@

The robot controller (with the help of a task planner) decides what action to
execute next, and who should perform it from the human or the robot (or both in
case of a joint action), and finally controls or monitors its execution. The
operation continues until the goal is achieved, is declared unachievable or is
abandoned by the human~\cite{Klein2004}.

@@

Such abilities must be designed and implemented in a generic way and
must provide several levels of parametrization such as they adapt to
various environments, different tasks and variable levels of engagement of the
robot, ranging from teammate behaviour to assistant or pro-active helper.

@@

The next section introduces the architecture as a whole, as well as the
knowledge model that we have developed for our robots. Then,
section~\ref{sec:impl} gives some insights on each of the main components of the
architecture, and tries to highlight their significance for artificial
intelligence. Section~\ref{sec:expe} presents two studies that illustrate in a
practical way what can be currently achieved with our robots.
Section~\ref{sect|conclusion} finally restate our main contributions and discuss
the key challenges that human-robot interaction brings to artificial
intelligence.

@@

  %%% Relations between components
  \path (shary.340) edge [<->, bend left] node[label] {motion plan \\ requests} (mhp);
  \path (shary.west) edge [<->, bend right] node[label] {shared \\ plans} (hatp);
  \path (hatp) edge [<->, bend right] node[label] {world model and \\ agents beliefs} (oro.170);
  \path (dialogs) edge [<->, bend left] node[label] {natural language \\ grounding} (oro.190);
  \path (spark.100) edge [->, bend right] node[label] {symbolic \\ facts} (oro);
  \path (spark.5) edge [->, bend right] node[label] {environment\\model} (mhp);
  \path (shary) edge [<->, bend left] node[label] {events, \\ world model and \\ agents beliefs} (oro);
  \path (shary) edge [<->, bend left] node[label] {action monitoring \\ and management of \\ position hypotheses} (spark);
  \path (lowlevel) edge [->] (spark);
  \path (lowlevel.east) edge [<-, bend right=80, looseness=1.2] node[label] {atomic\\actions} (shary.east);

@@

In particular, dealing with the rich semantics of natural interaction with
humans has been one major focus. To this end, we have been deploying
over the last years explicit knowledge representation and manipulation in the
deliberative layer of the robots as a {\it lingua franca} between our
components.

@@

To some extent, this architecture moves away from classical layered approaches
found in robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}.
Interactions between components at the deliberative level are mostly
bidirectional and we do not introduce layers of abstraction amongst software
components\footnote{We do have lower-level modules to execute actions or manage
sensors, but all cognition-related modules live in the same global deliberative
layer.}. Dialogue processing, for instance, illustrates this structure. This
component does not purely act as an alternative sensing modality that would be
fed to a distinct deliberative component. On the contrary, it actively queries
the knowledge base to interpret and disambiguate natural language, and in that
regards, it is fully integrated to the deliberative process itself (we explain
natural language processing at section~\ref{sect|com}).

@@

At run-time, the knowledge available to the robot at a given time comes from
three sources. {\it A priori} knowledge is stored in an ontology (the {\sc
OpenRobots} ontology, discussed below) and loaded at start-up. This static source covers the
\emph{common-sense} knowledge of the robot, and optionally some
scenario-specific knowledge (for instance, about objects that are to be
manipulated). The second part of the knowledge is acquired at run-time from
perception, interaction and planning. The next sections go into details of these
processes.  Lastly, the third source of symbolic statements is the reasoner
itself.

@@

    \node [node distance=1, below right=of thing] (tthing) {temporal thing} edge (thing);
        \node [below=of tthing] (tte) {thing with temporal extent} edge (tthing);
            \node [below right=of tte] {time interval} edge (tte);
            \node [below=1 of tte] (sit) {situation} edge (tte);
                \node [below right=of sit] (evt) {event} edge (sit);
                    \node [below right=of evt] (act) {action} edge (evt);
                        \node [below=of act] {purposeful action} edge (act);
                \node [below left=of sit] {static situation} edge (sit);

@@

Aligning the upper part of the ontology on {\sc OpenCyc} (as done by other
knowledge representation systems, like {\sc KnowRob}~\cite{Tenorth2009a} or PEIS
K\&R~\cite{Daoutis2009}) has multiple advantages. First the design of this part
of the ontology is generally difficult: it pertains to abstract concepts whose
mutual relations comes to philosophical debates. The upper taxonomy of {\sc
OpenCyc} represents a relative consensus, at least within the semantic Web
community. Then, because it is a well established project with numerous links to
other on-line databases (like Wikipedia or WordNet), the reuse of key {\sc
OpenCyc} concepts ensures to a certain extent that the knowledge stored by the
robot can be shared or extended with well-defined semantics. A good example is
the concept of \emph{Object}: in everyday conversation, an object is a
relatively small physical thing, that can be typically manipulated. Normally, a
human is not considered as an object. In {\sc Cyc}, an object has a more precise
definition: it is something \emph{partially tangible}. This includes obviously
the humans, and actually many other entities that would not be commonly said to
be objects (the Earth for instance). Thus the importance of relying on
well-defined and standard semantics to exchange informations between artificial
systems in a non-ambiguous way.

@@

The class \concept{PurposefulAction} is the superset of all the actions that are
purposefully performed by the robot (or another agent). Actual actions (\ie
subclasses of \concept{PurposefulAction} like \concept{Give} or
\concept{LookAt}) are not initially asserted in the common-sense ontology. They
are instead added at run-time by the execution controller (in link with the
symbolic task planner) and the natural language processor based on what the
robot is actually able to perform and/or to interpret in the current context
(\ie the current robot configuration and the actions required by the scenario).
The set of actions that the robot can interpret typically corresponds to the
planning domain in use: the robot can only process the \concept{GoTo} action if
the corresponding task description with its pre- and post-conditions is
available to the symbolic task planner.

@@

\emph{Grounding} (also called \emph{anchoring} when specifically referring to
the building of links between \emph{percepts} and \emph{physical
objects}~\cite{Coradeschi2003}) is the task consisting in building and
maintaining a bi-directional link between sub-symbolic representations (sensors
data, low-level actuation) and symbolic representations that can be manipulated
and reasoned about~\cite{Harnad1990}. This represents an important cognitive
skill, in particular in the human-robot interaction context: in that case, the
link that the robot has to established between its percepts and its symbol must
\emph{model} to a large extent the human representations in order to effectively
support communication.

\emph{Symbol grounding} connects hence the knowledge model to the perception and
actuation capabilities of the robot, and our architecture adopts here also a
bidirectional approach: \emph{bottom-up} and \emph{top-down}.  The different
components that we have mentioned so far provide a bottom-up grounding process:
geometric reasoning and dialogue processing modules constantly build and push
new symbolic contents about the world to the knowledge base where it becomes
accessible to decisional layers. In parallel, the knowledge base relies on
reasoning in a top-down way to produce new facts, \ie new beliefs that may lead
to actual decision making and physical actions.

@@

\begin{inparaenum} In our context, we call \emph{cognitive skills} the
deliberative behaviours that are \item \emph{stateful} (keeping track of
previous states is typically required for the component to perform adequately),
\item \emph{amodal} in that the skill is not inherently bound to a specific
perception of actuation modality, \item manipulate \emph{explicit semantics},
typically by the mean of symbolic reasoning, \item operate at the
    \emph{human-level}, \ie are legible to the humans, typically by acting at
similar levels of abstraction.\end{inparaenum}

@@

\subsubsection{Symbolic Reasoning}
\label{reasoning}

@@

We rely however on this mechanism in certain cases: for instance, some modules
like the natural language processor use the {\it short term} memory profile to
mark for a few seconds important concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory). We use
this feature to trace certain knowledge-related processes. On the other hand,
our perception layer does not make use of this mechanism: as described in the
next section, the environment model of the robot is continuously updated and the
derived symbolic knowledge is therefore transient.

@@

\begin{figure}
\centering
\resizebox{0.8\textwidth}{!}{%
\begin{tikzpicture}[
        >=latex,
        box/.style={draw,rectangle, dotted, minimum width=#1, minimum height=3.8cm},
        box/.default={4cm},
        every edge/.style={draw, ultra thick},
        every node/.style={align=center}]

@@

    \node[draw=none] at (0,-1.5) {\includegraphics[width=1.\columnwidth]{spark-oro-notext.pdf}};

    \node[box=16cm, minimum height=5.3cm, solid, thick, rounded corners] at (0,1.8) (oro) {};
    \node at (0,4.2) {\sc \Large oro-server};

@@

    \node at (-3,-0.4) {{\it OWL store:} OpenJena};
    \node at (3,-0.4) {{\it Reasoner:} Pellet};

@@

    \node[box=18cm, minimum height=4.5cm, dotted, thick, rounded corners] at (0,-4.2) (spark) {};
    \path (spark.north) edge[->] node[align=left,anchor=west] {Symbolic\\facts} (oro.south);

    \node[anchor=north] at (-6.5,-5.3) {a) Agent-independent\\locations};
    \node[anchor=north] at (-2.1,-5.3) {b) Agent-dependent\\placements};
    \node[anchor=north] at (1.6,-5.3) {c) Visibility};
    \node[anchor=north] at (6.5,-5.3) {d) Reachability};

@@

{\sc Spark} embeds an \emph{amodal}~\cite{Mavridis2006} geometric model of the
environment that serves both as basis for the fusion of the perception
modalities and as bridge with the symbolic layer. This geometric model is made
of CAD 3D models of the objects, furnitures and robots, and full body, rigged
models of the humans, and it is updated at run-time by the robot based on its
sensors (in most of our studies, objects are identified and localised through 2D
fiducial markers, while humans are tracked with Kinect-like devices, optionally
assisted by motion capture to accurately track the head motion, which is
required to compute what the human is looking at).

@@

{\sc Spark} is designed to run and continuously update the knowledge base (in
our experiments, it runs at about 10Hz). At each step, it re-computes spatial
relationships and affordances for the whole scene, and send the delta (new
relations and relations that do not hold anymore) to the knowledge base. This
approach may raise scalability concerns (we did not however hit major
performance issues in our constrained scenarios), but greatly simplifies the
management of the \emph{dynamics} of the knowledge (\emph{When do I discard
outdated knowledge?  When do I update it?}). Since it is equivalent to a reset
of the reasoner domain, it also hides issues linked to non-monotonic reasoning
(see~\cite{McCarthy2007} for a discussion on that question).

@@

\item \emph{Sees}: this relation describes what the agent can see, \ie what is
    within its field of view (FOV). In our current implementation, this
    affordance is computed by dynamically placing an OpenGL camera at the
    location of the eyes and running occlusion checks from it.  In
    figure~\ref{fig|spark-oro}\emph{c} the field of view of a person is
    illustrated with a grey cone (broader one). While he is able to see the two
    small boxes on the table in front of him, the big box on his right is out of
    his FOV, and therefore, he is not able to see it.

@@

\fxfatal{Rephrase or remove}
It is sometimes difficult or even impossible to see and/or track an object in
certain states. This happens, for instance, when the object is hidden in a
container, when the robot holds it in its gripper, and more generally in any
state in which it is hidden by something else. {\sc Spark} models the possible
symbolic states of objects (whether the object is on a furniture, in an agent
hand, in a container, etc.): according to the robot perception of what has
happened since the object was last seen, the robot builds a belief of the
current possible state with associated probability model.

@@

The main weakness of our current implementation relates to the way semantics are
conveyed: because we use 2D markers for artifacts, we avoid issues related to object
segmentation and recognition. Each object has a unique identifier,
which enables us to load a proper CAD model suitable for 3D spatial reasoning
and prevent recognition mistakes in the knowledge base. While {\sc Spark}
algorithms themselves are ignorant of the input sources, and would work equally
well with a full object recognition stack, we do not have tackled it yet.

Besides, temporal reasoning (essential for accurate action recognition for
instance) is not properly addressed in our stack. Temporal reasoning is used
only locally, and does not allow for tracking of long sequences or global events.

@@

In our experience, natural language processing is one of the fields of
human-robot interaction for which the introduction of the semantic layer has
been most beneficial.  In~\cite{Lemaignan2011a} we explain the techniques and the
tool called {\sc Dialogs} that we have developed for natural English language
parsing and grounding, along with verbalisation and (minimalist) dialogue
management.

@@

The system supports quantification (``give me \{a | the | some | all | any |
...\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also permits knowledge extension by learning new semantic
structures (for instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
after checking for possible contradictions with existing knowledge),
interprets common temporal and place adverbs (like \emph{above} or
\emph{tomorrow}) and translates to a certain extent \emph{states} (``I'm
tired'') into \emph{experiences} (\stmt{HUMAN experiences state\_1, state\_1
hasFeature tired}). We kindly refer the readers to~\cite{Lemaignan2011a} for a
complete discussion of these capabilities and the related algorithms.

\subsubsection{Multi-Modality}

@@

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{human1 desires
?action, ?action type Take, ?action actsOnObject ?obj, ?obj type Toy}, we could trigger a
behaviour when the human expresses (through dialogue, gestures...) that he
wants a toy.

@@

Manipulation and navigation actions listed in table~\ref{table|pyrobots_actions}
rely on a dedicated 3D motion planner ({\sc Mhp}, depicted in
Figure~\ref{fig|archi}): robot placement and end-effectors trajectories are
computed on-line to allow object manipulation that take into account task
specific constraints and human postures, abilities and preferences.
We kindly refer the interested reader to~\cite{Sisbot2008, Mainprice2011,
Pandey2011} for details of these techniques.

@@

This leads to a tight coupling between the symbolic and the geometric realms:
while AI at its origins was mostly a matter of symbolic models, it has been
since recognised that not only the mind is not a purely abstract system,
disconnected from the physical world, but even more, cognition fundamentally
relies on its relation to the physical world (so-called \emph{embodied
cognition}). Varela~\cite{Varela1992} is one of the main discoverers of these
mechanisms, and coined the concept of \emph{enactivism} as the theoretical
framework that study the links between cognition, embodiment and actions.

@@

Grounding also occurs during verbal and non-verbal communication. The {\sc
Dialogs} module grounds new concepts introduced by the speaker by asking questions
until it can attach the new concept to concepts already present in the
knowledge pool (typically, if I ask the robot ``bring me a platypus'', the
robot asks ``what is a platypus''. I may answer ``a mammal'' and the robot will
ask ``what is a mammal?'' and so on until it anchors the new concepts to ones
it already knows). Non verbal interaction (like gestures) is also partially
grounded: we have presented how pointing, for instance, is used by the robot to
ground \emph{``this''} or \emph{``that''} to the pointed artifact.

@@

As a whole, the components that we have introduced so far build a
knowledge-oriented architecture: knowledge is explicitly stored in one central
and consistent repository of facts, accessible for all modules. It relies on a
strict formalism (OWL statements), with a well defined vocabulary
(stated in the common-sense ontology). These first two points lead to a
\emph{loosely-coupled architecture} where modules can be removed or replaced
easily by other ones as long as they share the same semantics (modules are
defined by the knowledge they produce).

@@

And finally, this architecture allows for the combination of very different
knowledge sources in a single homogeneous environment, bringing mutual
benefits to components. For instance, the dialogue processing module can
run without any situation assessment, but its disambiguation routines
can transparently benefit from it when available (since richer symbolic
descriptions of objects are then available).
